Note on: Connections between Smoothing Splines and Laplacian smoothing

### 2021-05-04 notes/misc/splines_graph_laplacians_connections.tex
Partway through updating note to include results on Laplacian smoothing out of sample.

Recall the definition of an \emph{order-$m$ smoothing spline}. For observations $(x_1,y_1),\ldots,(x_n,y_n)$, design points $x_i$ each belonging to a domain $\Omega \subset \Rd$, 
\begin{equation}
\label{eqn:smoothing_spline}
\wh{g} = \argmin_{g \in G} \frac{1}{n} \sum_{i = 1}^{n} (y_i - g(x_i))^2 + \lambda |g|_{m}^2;
\end{equation}
here we use $|g|_{m}$ to denote the $H^m(\Omega)$, and $G$ to be an appropriately defined function class. The standard choice is $G = H^m(\Omega)$, in which case we must also choose $m > d/2$ to make~\eqref{eqn:smoothing_spline} be well-defined.

The goal of this note is to review properties of order-$m$ smoothing splines, and see if/when graph Laplacian smoothing possesses analogous properties.

\section{Well-posedness of smoothing splines}
\textcolor{red}{(TODO)}

\section{Optimality of smoothing splines}
Suppose $x_1,\ldots,x_n$ are sampled independently from a \textcolor{red}{sufficiently regular} density $p$ on $\Omega$, and that
\begin{equation}
\label{model:regression}
y_i = f_0(x_i) + \varepsilon_i,
\end{equation}
where $\varepsilon_i \sim N(0,1)$ are i.i.d. Gaussian noise that are also independent of $x_i$. Assume that $f_0 \in H^s(\Omega)$. What can we say about mean squared error $\|\wh{g} - g\|_P^2$?

\textcolor{red}{(TODO)}:
\begin{itemize}
	\item Smoothing splines are optimal when $2m > s$. Is there a lower bound on $s$ relative to $m$ or $d$?
	\item When $m < s < 2s$, the optimality requires some boundary conditions on $f_0$. Detail these boundary conditions and explain intuition behind why they are required. Compare with the boundary conditions required for orthogonal series/non-parametric least squares estimators to be optimal.
 	\item Detail various techniques for establishing optimality. Refer to \textcolor{red}{(van der Geer, Nussbaum, Cox)}. Review pros and cons of each technique.
\end{itemize}

\section{Low-smoothness regime}
We now introduce a \emph{restricted smoothing spline}, in order to compute estimates for $m < d/2$. Suppose we take $\Phi_N \subseteq G$ to be a finite set of $L^2(P)$-orthogonal low-frequency functions; formally $\Phi_N = \{\phi_1,\ldots,\phi_N\}$, where each $\phi_k$ satisfies the Laplacian eigenfunction equation,
\begin{equation}
\label{eqn:laplacian_eigen}
\Delta_P \phi_k = \lambda_k \phi_k,\quad \frac{\partial}{\partial {\bf n}}\phi_k = 0.
\end{equation}
Let $F_N = \mathrm{span}\{\Phi_N\}$, and consider
\begin{equation}
\label{eqn:restricted_smoothing_spline}
\wh{g}_{\Phi} = \argmin_{g \in F_N} \frac{1}{n} \sum_{i = 1}^{n} (y_i - g(x_i))^2 + \lambda |g|_{m}^2.
\end{equation}
The restricted smoothing spline~\eqref{eqn:smoothing_spline} is well-posed, and its solution $\wh{g}_{\Phi}$ possesses a relatively simple analytic solution. It may be totally impossible to exactly compute because it is difficult to exactly solve~\eqref{eqn:laplacian_eigen}, but that is besides the point. Now we analyze its statistical properties. To do so, we reduce the analysis to that of a related estimator in the Normal means/Gaussian sequence model.

\textcolor{red}{(TODO)}
\begin{itemize}
	\item Refer to the results of \textcolor{blue}{(Birge and Massart)} for empirical risk minimization (ERM) over normed balls in H\"{o}lder spaces. Resolve why Birge and Massart find a phase transition at $m = 1, d = 2$, which is not at the $4m = d$ boundary; I suspect it is because Birge and Massart find the point at which the empirical process diverges.
\end{itemize}

\subsection{Gaussian sequence model}
In the Gaussian sequence model, for each $i \in \mathbb{N}$ one observes
\begin{equation}
\label{model:gaussian_sequence_model}
Y_i = \theta_i^{\ast} + \frac{1}{\sqrt{n}}\varepsilon_i,\quad \varepsilon_i \sim N(0,1).
\end{equation}
Supposes the sequence $\theta^{\ast}$ possesses some ``smoothness''. In particular, suppose $\theta^{\ast}$ belongs to the \emph{order-$s$ Sobolev ellipsoid} of radius $M$, $\Theta^s = \{\theta: |\theta|_s \leq M\}$, where for $s \in \mathbb{N}$ the Sobolev seminorm $|\theta|_s$ is defined by
\begin{equation*}
|\theta|_s^2 := \sum_{j = 1}^{\infty} \theta_j^2 j^{2s/d}.
\end{equation*}
Consider the \emph{sequence smoothing spline} estimator, defined for $N \in \mathbb{N}$ as
\begin{equation}
\label{eqn:sequence_smoothing_spline}
\wh{\theta}_{N} := \argmin_{\theta \in \ell^2(N)} |Y - \theta|_{0} + \lambda |\theta|_m^2, \quad \wh{\theta} := \lim_{N \to \infty} \wh{\theta}_{N}.
\end{equation}
Note that the sequence space estimator $\wh{\theta}$ is perfectly well-posed even though its analogous function space estimator is not. In fact, the sequence space estimator has the explicit solution
\begin{equation}
\wh{\theta}_{j} = Y_j[1 + \lambda j^{2m/d}]^{-1},
\end{equation}
showing that it is a shrinkage estimator.

@@

	\item Relate the sampling model~\eqref{model:gaussian_sequence_model} to~\eqref{model:regression}, and likewise relate the estimator~\eqref{eqn:sequence_smoothing_spline} to~\eqref{eqn:restricted_smoothing_spline}. Refer to \textcolor{blue}{(Tsybakov, Johnstone)}.
	\item Show that in the Gaussian sequence model, if $m > \max\{d/4,s/2\}$ then~\eqref{eqn:sequence_smoothing_spline} achieves optimal $\ell_2$ loss when $N = \infty$.

@@

On the other hand, whenever $m \leq d/4$ meaningful restrictions on the size of $N$ kick in.

@@

	\item When $r = 1$---corresponding to the first-order Sobolev norm--- Lemma~\ref{lem:smoothness_of_sequence_smoothing_splines} tells us that the sequence smoothing spline $\wh{\theta}$ has infinite $|\cdot|_1$ norm when $d \geq 2$. Remarkably, when $d = 2$ or $d = 3$, $\wh{\theta}$ continues to be optimal despite being an incredibly ``wiggly'' estimator of a ``smooth'' sequence. Note that this forecloses upon the possibility of analyzing $\wh{\theta}$ using uniform convergence over $\Theta^1(M)$, since $\wh{\theta}$ does not belong to this class for any finite $M$.
	\item When $r = 0$---corresponding to the zeroth-order Sobolev norm, or in other words the $\ell^2$-norm---Lemma~\ref{lem:smoothness_of_sequence_smoothing_splines} tells us that the sequence smoothing spline $\wh{\theta}$ has infinite $\Theta^0$ norm when $d \geq 4$. Clearly, since the regression function $f_0$ has finite $\Theta^0$ norm, and the estimator $\wh{f}$ has infinite $\Theta^0$ norm, it follows from the triangle inequality that $|\wh{f} - f|_0 = \infty$. 

@@

\textcolor{red}{(TODO)}
\begin{itemize}
	\item Do an analogous calculation for the $\Theta^{1,1}$ case, with the estimator
	\begin{equation*}
	\wt{\theta}_{N} = \argmin_{\theta \in \ell^2(N)} |Y - \theta|_0^2 + \lambda |\theta|_{1,1}^2.
	\end{equation*}
	 Confirm your expectation: that when error measured in $|\cdot|_0$ norm and $d > 1$, this estimator is optimal when $N = n$, but not when $N = \infty$. This would give some evidence that the graph is saving the TV denoiser when $d > 1$.
	 \item Figure out other smoothness properties of $\wh{\theta}_N$, for instance ``H\"{o}lder'' smoothness.
	 \item Ponder what implication the lack-of-smoothness of sequence smoothing splines has for other problems (besides estimation with error measured in $|\cdot|_0$ norm).
\end{itemize}

\section{Relationship to Graph Laplacian methods}

@@

>>>>

Recall the definition of an \emph{order-$m$ smoothing spline}. For observations $(x_1,y_1),\ldots,(x_n,y_n)$, design points $x_i$ each belonging to a open set with smooth boundary $\Omega \subset \Rd$, 
\begin{equation}
\label{eqn:smoothing_spline}
\wh{g} = \argmin_{g \in H^1(\Omega)} \frac{1}{n} \sum_{i = 1}^{n} (y_i - g(x_i))^2 + \lambda |g|_{m}^2;
\end{equation}
here we use $|g|_{m}$ to denote a Hilbert-Sobolev semi-norm on functions $H^m(\Omega)$, and we insist that $m > d/2$ to make~\eqref{eqn:smoothing_spline} be well-defined.

The goal of this note is to review properties of the order-$m$ smoothing spline, to see if/when graph Laplacian smoothing possesses analogous properties, and to see if/when graph Laplacian smoothing possesses different properties.

\section{Gaussian sequence model}
All calculations are easy in the Gaussian sequence model, and we begin there. Much of what we do is extracted from \textcolor{blue}{(Johnstone)}. In the Gaussian sequence model, for each $i \in \mathbb{N}$ one observes
\begin{equation}
\label{model:gaussian_sequence_model}
Y_i = \theta_i^{\ast} + \frac{1}{\sqrt{n}}\varepsilon_i,\quad \varepsilon_i \sim N(0,1).
\end{equation}
Consider the order-$m$ \emph{sequence smoothing spline} estimator,
\begin{equation}
\label{eqn:sequence_smoothing_spline}
\wh{\theta} := \argmin_{\theta \in \ell^2(\mathbb{N})} |Y - \theta|_{0} + \lambda |\theta|_m^2.
\end{equation}
where for $m \in \mathbb{N}$ the Sobolev seminorm $|\theta|_m$ is defined by
\begin{equation*}
|\theta|_m^2 := \sum_{j = 1}^{\infty} \theta_j^2 j^{2m/d}.
\end{equation*}
Note that the order-$m$ sequence smoothing spline $\wh{\theta}$ is perfectly well-posed when $m < d/2$, even though its analogous function space estimator is not. In fact, the sequence space estimator has the explicit solution
\begin{equation}
\wh{\theta}_{j} = Y_j[1 + \lambda j^{2m/d}]^{-1},
\end{equation}
showing that it is a shrinkage estimator.

Now we analyze the minimax risk of $\wh{\theta}$ over Sobolev ellipsoids. That is, we upper bound the following quantity,
\sup_{\theta^{\ast} \in \Theta^s} \Ebb\bigl[|\wh{\theta} - \theta^{\star}|_{0}^2\bigr]
where $\Theta^s = \{\theta: |\theta|_s \leq M\}$ is the \emph{order-$s$ Sobolev ellipsoid}. Note that we do not assume $s = m$. Note also that $|\cdot|_0$ is the usual norm over $\ell^2(\mathbb{N})$. Finally, a word on terminology: we will use the term ``order-$m$ smoothness'' to refer to the size of the squared norm $|\theta|_m^2$. Of course, these are sequences, and the notion of ``smooth'' sequences is not a totally natural concept. We use this wording regardless, to unify our discussion of Gaussian sequence and non-parametric regression models.

@@

	\item Show that in the Gaussian sequence model, if $m > \max\{d/4,s/2\}$ then~\eqref{eqn:sequence_smoothing_spline} achieves optimal risk.

@@

On the other hand, whenever $m \leq d/4$ the estimator $\wh{\theta}$ is grossly sub-optimal. 

@@

	\item Consider $r = 1$, which corresponds to the first-order Sobolev norm. Lemma~\ref{lem:smoothness_of_sequence_smoothing_splines} tells us that the sequence smoothing spline $\wh{\theta}$ has infinite first-order Sobolev norm in expectation, $\Ebb|\wh{\theta}|_1 = \infty$, whenever $d \geq 2$. Remarkably, when $d = 2$ or $d = 3$, $\wh{\theta}$ continues to have optimal risk despite being an incredibly ``wiggly'' estimator of a ``smooth'' sequence. Note that this forecloses on the possibility of analyzing $\wh{\theta}$ using uniform convergence over $\Theta^1(M)$, since $\wh{\theta}$ does not belong to this class for any finite $M$.
	\item Consider $r = 0$, which corresponds to the zeroth-order Sobolev norm, or in other words the $\ell^2$-norm. Lemma~\ref{lem:smoothness_of_sequence_smoothing_splines} tells us that the sequence smoothing spline $\wh{\theta}$ has infinite $\Theta^0$ norm when $d \geq 4$. Clearly, since the regression function $f_0$ has finite $\Theta^0$ norm, and the estimator $\wh{f}$ has infinite $\Theta^0$ norm, it follows from the triangle inequality that $|\wh{f} - f|_0 = \infty$. 

@@

\textcolor{red}{(TODO)}
\begin{itemize}
	\item Do an analogous calculation for empirical risk minimization over $\Theta^1(M)$. Intuition, both from the above calculations and from \textcolor{blue}{(Birge and Massart)}, suggest that there is a phase transition at $d = 2$.  
	\item Do an analogous calculation for the $\Theta^{1,1}$ case, with the estimator
	\begin{equation*}
	\wt{\theta}_{N} = \argmin_{\theta \in \ell^2(N)} |Y - \theta|_0^2 + \lambda |\theta|_{1,1}^2.
	\end{equation*}
	Confirm your expectation: that when error measured in $|\cdot|_0$ norm and $d > 1$, this estimator is optimal when $N = n$, but not when $N = \infty$. This would give some evidence that the graph is saving the TV denoiser when $d > 1$.
	\item Figure out other smoothness properties of $\wh{\theta}_N$, for instance ``H\"{o}lder'' smoothness.
	\item Ponder what implication the lack-of-smoothness of sequence smoothing splines has for other problems (besides estimation with error measured in $|\cdot|_0$ norm).
\end{itemize}

\section{High-smoothness regime: optimality of smoothing splines}
Suppose $x_1,\ldots,x_n$ are~\textcolor{red}{(well-spaced)} in $\Omega$, and that
\begin{equation}
\label{model:regression}
y_i = f_0(x_i) + \varepsilon_i,
\end{equation}
where $\varepsilon_i \sim N(0,1)$ are i.i.d. Gaussian noise that are also independent of $x_i$. Assume that $f_0 \in H^s(\Omega)$. What can we say about in-sample mean squared error $\|\wh{g} - g\|_n^2$? What can we say about the out-of-sample mean squared error $\|\wh{g} - g\|_{L^2(\mc{X})}^2$?

\textcolor{red}{(TODO)}:
\begin{itemize}
	\item Necessary conditions for the order-$m$ smoothing spline to be optimal are that $m > s/2$ and $m > d/2$. Are these sufficient conditions---at least, on the indices $(m,s,d)$---or is there a lower bound on $s$ relative to $m$ or $d$?
	\item When $s/2 < m < s$, the aforementioned optimality properties require that $f_0$ satisfy some boundary conditions. Detail these boundary conditions and explain why they are necessary. Compare with the boundary conditions required for orthogonal series/non-parametric least squares estimators to be optimal.
 	\item Detail various techniques for establishing optimality. Refer to \textcolor{red}{(van der Geer, Nussbaum, Cox)}. Review pros and cons of each technique.
\end{itemize}

\section{Low-smoothness regime: optimality of restricted smoothing splines}
We now introduce the order-$m$ \emph{restricted smoothing spline}, an alternative to the order-$m$ smoothing spline which makes sense when $m < d/2$. To define the order-m restricted smoothing spline, we first need to review the \emph{weighted Laplace-Beltrami operator} $\Delta_P$, and an associated eigenvector problem. For this purpose, we now introduce our formal assumptions on the design distribution $P$. Let $\nu$ denote the $d$-dimensional Lebesgue measure.
\begin{assumption}[Design distribution $P$.]
	\label{asmp:design}
	The design distribution $P$ is supported on an open, bounded set $\Omega \subseteq \Rd$, which has $C^{\infty}$ boundary $\partial \Omega$. The distribution $P$ admits a $C^{\infty}(\Omega)$ density $p$ with respect to Lebesgue measure $\nu$, which is bounded away from $0$ and $\infty$, meaning that for some $p_{\min} < 1 < p_{\max}$
	\begin{equation*}
	0 \leq p_{\min} \leq p(x) \leq p_{\max} < \infty,\quad\textrm{for all $x \in \Omega$.}
	\end{equation*}
\end{assumption} 
We note that the assumptions in~\eqref{asmp:design} are in a certain sense very strong, and quite likely overkill for any particular value of $s$ and $d$. The minimum conditions required on $P$ in order to have the operator $\Delta_P$ and its eigenvectors be ``sufficiently regular'' for statistical purposes is certainly of interest, but not the point of this note.

If $P$ satisfies Assumption~\ref{asmp:design}, the weighted Laplace-Beltrami operator $\Delta_P$ is defined as 
\begin{equation*}
\Delta_Pf := \frac{-1}{p} \mathrm{div}(p^2 \nabla f)~~\textrm{in $\Omega$}, \quad \frac{\partial f}{\partial {\bf n}} = 0~~\textrm{on $\partial \Omega$.}
\end{equation*}
(The second part of the statement above should interpreted in the usual way, to mean that $\Delta_P$ is an operator acting on those functions $f \in L^2(P)$ that satisfy Neumann boundary conditions). We see that $\Delta_P$ is a second-order symmetric differential operator, and is self-adjoint in $L^2(P)$; the assumptions on $p$ further imply that it is uniformly elliptic. As such it admits a discrete spectrum, with eigenvalues $0 < \lambda_1 \leq \lambda_2 \leq \cdots$, and corresponding eigenvectors $\psi_1,\psi_2,\ldots$ which form an orthonormal basis of $L^2(P)$. Standard theory (e.g. Theorem 6, Chapter 6.3 of \textcolor{blue}{(Evans)}) implies that each eigenvector $\psi_j \in C^{\infty}(\wb{\Omega})$.

Let $N$ be a positive integer. We define the order-$m$ restricted smoothing spline to be the penalized least squares estimator over the linear subspace  $\Phi_N = \mathrm{span}\{\psi_j: j \leq N\}$ of $L^2(P)$. Formally speaking,
\begin{equation}
\label{eqn:restricted_smoothing_spline}
\wh{g}_{N} = \argmin_{g \in \Phi_N} \frac{1}{n} \sum_{i = 1}^{n} (y_i - g(x_i))^2 + \lambda |g|_{m}^2.
\end{equation}
The order-$m$ restricted smoothing spline defined in~\eqref{eqn:restricted_smoothing_spline} is well-posed. Let $\Gamma \in \Reals^{N \times N}$ be the matrix with entries $\Gamma_{ij} = \dotp{\psi_i}{\psi_j}_{m}$. For a given $\alpha \in \Reals^N$, define $g_{\alpha} = \sum_{i = j}^{N} \alpha_j \psi_j$. Then
\begin{equation*}
\wh{g}_{N} = \argmin_{\alpha \in \Reals^n} \frac{1}{n} \sum_{i = 1}^{n} (y_i - g_{\alpha}(x_i))^2 + \lambda \alpha^T \Gamma \alpha.
\end{equation*}
This is just a generalized ridge problem, and can be solved either exactly, or approximately by an iterative procedure.

Unfortunately, it may be totally impossible to actually compute $\wh{g}_N$, because in order to do so we need access to the eigenfunctions $\psi_j$, which in turn requires knowing $P$. We ignore this problem for the moment, but return to it later when we introduce graph Laplacian smoothing methods. For now, we assume that we can compute the estimator defined by~\eqref{eqn:restricted_smoothing_spline}, and concentrate on analyzing its statistical properties.

\textcolor{red}{(TODO)}
\begin{itemize}
	\item Analyze the statistical properties of the estimator defined in~\eqref{eqn:restricted_smoothing_spline}. Note that you are no longer clear about whether or not these are the same as the properties of the sequence model.
\end{itemize}

\section{Supervised graph Laplacian smoothing}

@@

\section{Semi-supervised graph Laplacian smoothing}

Suppose now that in addition to the \emph{labeled} data $(x_1,y_1),\ldots,(x_n,y_n)$, we observe $m$ \emph{unlabeled} design points $x_{n + 1},\ldots,x_{n + m}$. Let $N = n + m$, and let $L_{N,\varepsilon}$ denote the graph Laplacian over labeled and unlabeled points,
\begin{equation*}
\bigl(L_{N,\varepsilon}u\bigr)_i := \frac{1}{N \varepsilon^{d + 2}}\sum_{j = 1}^{N} (u_i - u_j) \eta\biggl(\frac{\|x_i - x_j\|}{\varepsilon}\biggr).
\end{equation*}
The order-$m$ graph Laplacian smoothing estimator $\wh{f}$ defined in \textcolor{red}{(?)} can be naturally generalized to a vector in $\Reals^N$ according to the following formulation,
\begin{equation}
\label{eqn:ssl_graph_laplacian_smoothing_0}
\wh{f}_N := \argmin_{f \in \Reals^N} \|Y - f\|_n^2 + \lambda \dotp{L_{N,\varepsilon}f}{f}_N.
\end{equation}
The estimator $\wh{f}$ has a closed-form solution, which can be simply written using the extension operator $E_{n \to N}: \Reals^{n} \to \Reals^N$ and its adjoint, the restriction operator $R_{N \to n}: \Reals^{N} \to \Reals^n$, which are defined by
\begin{equation*}
E_{n \to N}u := (u_1,\ldots,u_n,0,\ldots,0),\quad\textrm{and}\quad R_{N \to n}v := (v_1,\ldots,v_n).
\end{equation*}
Then the order-$m$ graph Laplacian smoothing estimator can be written as follows
\begin{equation}
\label{eqn:ssl_graph_laplacian_smoothing_1}
\wh{f}_N = (E_{n \to N} R_{N \to n} + \frac{\rho n}{N} L_{N,\varepsilon})^{-1} E_{n \to N}Y.
\end{equation}
The form of~\eqref{eqn:ssl_graph_laplacian_smoothing_1} is a compact way to represent the solution $\wh{f}$, but it is rather opaque. In Proposition~\ref{prop:ssl_graph_laplacian_smoothing} we give a second form of $\wh{f}$, which is less compact but more evocative. 
\begin{proposition}
	\label{prop:ssl_graph_laplacian_smoothing}
	Let $\wh{f}_N$ be the solution to~\eqref{eqn:ssl_graph_laplacian_smoothing_0}. Writing $\wh{f}_N = (\wh{f}_n,\wh{f}_m)$ for $\wh{f}_n \in \Reals^m$ and $\wh{f}_n \in \Reals^m$, we have the following.
	\begin{itemize}
		\item 
	\end{itemize}
\end{proposition}
